---
title: "Beta Mixture Model"
author: "Shannon T. Holloway"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: "hide"
params:
  cloud_path: "~/Library/CloudStorage/Box-Box/PROJECT 06023: MolClocks/"
  data_dir: "MolClock_Paper_1/1. Analytic Datasets/TCGA/"
  data_dir_lund: "MolClock_Paper_1/1. Analytic Datasets/Lund/"
  working_dir: "~/Desktop/FromDesktop_2bsorted/cibersort/"
  methyl_filename: "cohort1.methyl"
  clinical_filename: "cohort1.clinical"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("~/Desktop/FromDesktop_2bsorted/cibersort/R_codes/my_BetaMixture.R")
library(ggsci)
library(ggplot2)
```

```{r retrieve_CpG_site}
cpg_sites <- read.csv(paste0(params$working_dir, "balanced_CpGs_final_set.txt"),
                      header = FALSE)[, 1L]
```

## TCGA

# Beta Peak Locations {.tabset}

We have decided to use the BetaMixture tool and will show results only for the
measured beta.

<span style="color:red">Missing values are a problem for the clustering algorithm.
Fitting without these sites for now.</span>

<span style="color:red">0's and 1' are a problem for the clustering algorithm.
Removing those sites.</span>

## BetaMixture

Package available on R-Forge. Required modification of a testing criteria
to avoid floating point issue (`sum(x) == 1` vs `isTRUE(all.equal(sum(x), 1.0))`)

```{r TCGA_beta_mixture_model}
betaMixture_TCGA <- tryCatch(readRDS(paste0(params$cloud_path, params$data_dir,
                                            "tcga_beta_mixture_model.rds")),
                             error = function(e) { NULL })


if (is.null(betaMixture_TCGA)) {
  TCGA_clinical <- readRDS(paste0(params$cloud_path, params$data_dir,
                                  params$clinical_filename, ".rds"))
  
  analysis_tumors <- TCGA_clinical$in_analysis_dataset
  
  brca.methyl <- readRDS(paste0(params$cloud_path, params$data_dir,
                                params$methyl_filename, ".rds"))
  keep <- substr(colnames(brca.methyl), 1, 12L) %in% TCGA_clinical$submitter_id[analysis_tumors]
  
  brca.methyl <- brca.methyl[, keep]
  
  # we are fitting only the 500 CpG sites
  brca.methyl <- brca.methyl[rownames(brca.methyl) %in% cpg_sites, ]

  alpha <- matrix(0.0, ncol(brca.methyl), 3L)
  delta <- matrix(0.0, ncol(brca.methyl), 3L)
  z <- list()
  track_number_removed_TCGA <- numeric(ncol(brca.methyl))
  for (i in 1L:ncol(brca.methyl)) {
    
    are_na <- is.na(brca.methyl[, i])
    rm_extremes <- !are_na & {brca.methyl[, i] <= 0.98 & brca.methyl[, i] >= 0.02}
    track_number_removed_TCGA[i] <- sum(!rm_extremes)
    
    res <- tryCatch(suppressWarnings(BetaMixture_my(brca.methyl[rm_extremes, i], K = 3, 
                                                    forever = 1500,
                                                    debug = FALSE,
                                                    epsilon = 1e-3, 
                                                    relative = 1e-3)),
                    error = function(e) {
                      message(i, " ", sum(is.na(brca.methyl[, i])), " ", e$message)
                      stop("T")
                      })
    
    if (is.null(res)) next
    if (!res@converged) {
      message(i, " ", res@converged)
    }
    alpha[i, ] <- res@mle[1L, ]
    delta[i, ] <- res@mle[2L, ]
    z[[i]] <- res@Z
  }
  
  saveRDS(list(alpha = alpha, delta = delta, Z = z), 
          file = paste0(params$cloud_path, params$data_dir,
                        "tcga_beta_mixture_model.rds"))
  
  print(table(track_number_removed_TCGA))
}
```

## Lund Data

```{r get_lund_data}
lund.methyl <- read.csv(paste0(params$cloud_path, params$data_dir_lund, 
                                  "beta_values_FULL_balanced_CpGs.txt"), sep = "\t")
rownames(lund.methyl) <- lund.methyl[, 1L]
lund.methyl <- lund.methyl[, -1L]
```

```{r Lund_beta_mixture_model}
betaMixture_Lund<- tryCatch(readRDS(paste0(params$cloud_path, params$data_dir_lund,
                                              "lund_beta_mixture_model.rds")),
                               error = function(e) { NULL })

if (is.null(betaMixture_Lund)) {
  
  alpha <- matrix(0.0, ncol(lund.methyl), 3L)
  delta <- matrix(0.0, ncol(lund.methyl), 3L)
  z <- list()
  track_number_removed_Lund <- numeric(ncol(lund.methyl))
  for (i in 1L:ncol(lund.methyl)) {
    
    are_na <- is.na(lund.methyl[, i])
    rm_extremes <- !are_na & {lund.methyl[, i] <= 0.98 & lund.methyl[, i] >= 0.02}
    track_number_removed_Lund[i] <- sum(!rm_extremes)
    
    res <- tryCatch(suppressWarnings(BetaMixture_my(lund.methyl[rm_extremes, ], K = 3, 
                                                    forever = 1500,
                                                    debug = FALSE,
                                                    epsilon = 1e-3, 
                                                    relative = 1e-3)),
                    error = function(e) {
                      message(i, " ", sum(is.na(lund.methyl[, i])), " ", e$message)
                      stop("T")
                    })
  
    if (is.null(res)) next
    if (!res@converged) {
      message(i, " ", res@converged)
    }
    alpha[i, ] <- res@mle[1L, ]
    delta[i, ] <- res@mle[2L, ]
    z[[i]] <- res@Z
  }
  
  saveRDS(list(alpha = alpha, delta = delta, Z = z), 
          file = paste0(params$cloud_path, params$data_dir,
                        "lund_beta_mixture_model.rds"))
  
  print(table(track_number_removed_Lund))
}
```