---
title: "Beta Mixture Model"
author: "Shannon T. Holloway"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: "hide"
params:
  recreate_TCGA: TRUE # Recreate results for TCGA cohort
  recreate_Lund: TRUE # Recreate results for Lund cohort
  report_updates: TRUE # Report if the existing results files match the results generated here 
  # If they do, then output a new RDS file with the date and time in the filename
  outdir_name: 'Beta Peak Decomposition' # Output directory in external data storage
  TCGA_methyl_filename: "TCGA.methyl.Clock_revision" # Base name of TSV file with TCGA beta values
  Lund_methyl_filename: "beta_values_Clock_CpGs_revision_pureSamples" # Base name of TSV file with Lund beta values
  TCGA_clinical_filename: "TCGA.clinical" # Base name of RDS file with TCGA clinical data
  TCGA_output_filename: "tcga_beta_mixture_model_revision.rds" # Output file name of TCGA results
  Lund_output_filename: "lund_beta_mixture_model_revision.rds" # Output file name of Lund results
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("BetaMixture_0.5.2.R")
library(ggsci)
library(ggplot2)
library(jsonlite)

consts <- read_json(file.path(repo_dir, 'src', 'consts.json'))

# Import DNA methlation data table
# Set column and row names correctly
# Convert data to floats
importMethylTable <- function(methyl.path) {
    methyl <- read.table(methyl.path, sep='\t', header=F)
    names(methyl) <- methyl[1L, ]
    methyl <- methyl[-1L, ]
    rownames(methyl) <- methyl[, 1L]
    methyl <- methyl[, -1L]
    for (j in (1:ncol(methyl))) {
      methyl[, j] <- as.numeric(methyl[, j])
    }
    methyl
  }
```

## TCGA

# Beta Peak Locations {.tabset}

We have decided to use the BetaMixture tool and will show results only for the
measured beta.

<span style="color:red">Missing values are a problem for the clustering algorithm.
Fitting without these sites for now.</span>

<span style="color:red">0's and 1' are a problem for the clustering algorithm.
Removing those sites.</span>

## BetaMixture

Package available on R-Forge. Required modification of a testing criteria
to avoid floating point issue (`sum(x) == 1` vs `isTRUE(all.equal(sum(x), 1.0))`)

```{r TCGA_beta_mixture_model}
########################################################
outdirpath <- file.path(consts[['official_indir']], params$outdir_name, 'TCGA')
if (!dir.exists(outdirpath)) {
  dir.create(outdirpath, recursive = TRUE)
}
########################################################

if (params$recreate_TCGA) {
  cat("Rerunning TCGA Beta Decomposition\n")
  TCGA_clinical <- readRDS(file.path(consts['TCGA_datadir'],
                                  paste0(params$TCGA_clinical_filename, ".rds")))
  
  analysis_tumors <- TCGA_clinical$in_analysis_dataset
  
  TCGA.methyl <- importMethylTable(file.path(consts['TCGA_datadir'],
                                  paste0(params$TCGA_methyl_filename, ".tsv")))
  
  
  keep <- substr(colnames(TCGA.methyl), 1, 12L) %in% TCGA_clinical$submitter_id[analysis_tumors]
  
  TCGA.methyl <- TCGA.methyl[, keep]
  
  ########################################################  
  too_many_na_values <- colMeans(is.na(TCGA.methyl)) >= 0.05
  TCGA.methyl <- TCGA.methyl[, !too_many_na_values]
  cat(ncol(TCGA.methyl), " tumors; removed ", sum(too_many_na_values), "due to NA content", "\n")
  ########################################################  
  alpha <- matrix(0.0, ncol(TCGA.methyl), 3L, 
                  dimnames = list(colnames(TCGA.methyl), NULL))
  delta <- matrix(0.0, ncol(TCGA.methyl), 3L, 
                  dimnames = list(colnames(TCGA.methyl), NULL))
  z <- list()
  phi <- matrix(0.0, ncol(TCGA.methyl), 3L)
  track_number_removed_TCGA <- numeric(ncol(TCGA.methyl))
  track_number_extrema_TCGA <- numeric(ncol(TCGA.methyl))
  for (i in 1L:ncol(TCGA.methyl)) {
    
    are_na <- is.na(TCGA.methyl[, i])
    rm_extremes <- !are_na & {TCGA.methyl[, i] <= 0.98 & TCGA.methyl[, i] >= 0.02}
    track_number_removed_TCGA[i] <- sum(!rm_extremes)
    track_number_extrema_TCGA[i] <- sum(!are_na & {
      TCGA.methyl[, i] > 0.98 | TCGA.methyl[, i] < 0.02})
   
    res <- tryCatch(suppressWarnings(BetaMixture(TCGA.methyl[rm_extremes, i], K = 3, 
                                                 forever = 1500,
                                                 debug = FALSE,
                                                 epsilon = 1e-3, 
                                                 relative = 1e-3)),
                    error = function(e) {
                      message(i, " ", sum(is.na(TCGA.methyl[, i])), " ", e$message)
                      stop("T")
                      })
    hist(res, main = paste("Tumor ", colnames(TCGA.methyl)[i]))
    if (is.null(res)) next
    if (!res@converged) {
      message(i, " ", res@converged)
    }
    alpha[i, ] <- res@mle[1L, ]
    delta[i, ] <- res@mle[2L, ]
    z[[i]] <- res@Z
    phi[i, ] <- res@phi
  }
  names(z) <- colnames(TCGA.methyl)
  rownames(phi) <- colnames(TCGA.methyl)
  names(track_number_removed_TCGA) <- colnames(TCGA.methyl)
  
  ########################################################  
  ############# Saving results as R object ###############
  ########################################################
  TCGA_output_filepath <- file.path(outdirpath, params$TCGA_output_filename)
  
  # Collect old results
  if (params$report_updates) {
    ident.vec <- NULL
    if (file.exists(TCGA_output_filepath)) {
      prev.results <- readRDS(TCGA_output_filepath)
      
      # Compare all data objects
      identical.all <- all(c(
        identical(prev.results$alpha, alpha),
        identical(prev.results$delta, delta),
        identical(prev.results$Z, z),
        identical(prev.results$n_dropped_sites, track_number_removed_TCGA),
        identical(prev.results$phi, phi)
        ))
      
      if (identical.all) {
        print('Results DO match previously saved results, resaving anyway')
      } else{
        print('Results do not match previously saved results')
  
        TCGA_output_filepath <- paste(c(TCGA_output_filepath, as.character(Sys.time())), collapse='.')
        sprintf('Will not override, saving results instead as %s', TCGA_output_filepath)
      }
    }
  }
  
  # Save results in R data object just in case
  saveRDS(list(alpha = alpha, delta = delta, Z = z, 
               n_dropped_sites = track_number_removed_TCGA,
               phi = phi), 
          file = TCGA_output_filepath)
  
  ########################################################
  ########## Output results as readable tables ###########
  ########################################################
  write.table(alpha, file.path(outdirpath, 'alpha.txt'), quote=F)
  write.table(delta, file.path(outdirpath, 'delta.txt'), quote=F)
  write.table(phi, file.path(outdirpath, 'phi.txt'), quote=F)
  
  modes <- (alpha - 1)/(alpha + delta - 2)
  write.table(modes, file.path(outdirpath, 'modes.txt'), quote = F)
  ########################################################
  ########################################################
  
  print(table(track_number_removed_TCGA))
  print(table(track_number_extrema_TCGA))
  sprintf('Total # extreme values removed: %d', sum(track_number_extrema_TCGA))
}
```

## Lund Data

```{r Lund_beta_mixture_model}
########################################################
outdirpath <- file.path(consts[['official_indir']], params$outdir_name, 'Lund')
if (!dir.exists(outdirpath)) {
  dir.create(outdirpath, recursive = TRUE)
}
########################################################

if (params$recreate_Lund) {
  cat("Rerunning Lund Beta Decomposition\n")
  
  Lund.methyl <- importMethylTable(file.path(consts['Lund_datadir'],
                                  paste0(params$Lund_methyl_filename, ".tsv")))

  ########################################################  
  too_many_na_values <- colMeans(is.na(Lund.methyl)) >= 0.05
  Lund.methyl <- Lund.methyl[, !too_many_na_values]
  cat(ncol(Lund.methyl), " tumors; removed ", sum(too_many_na_values), "due to NA content", "\n")
  ########################################################  
  alpha <- matrix(0.0, ncol(Lund.methyl), 3L,
                  dimnames = list(colnames(Lund.methyl), NULL))
  delta <- matrix(0.0, ncol(Lund.methyl), 3L,
                  dimnames = list(colnames(Lund.methyl), NULL))
  z <- list()
  phi <- matrix(0.0, ncol(Lund.methyl), 3L)
  track_number_removed_Lund <- numeric(ncol(Lund.methyl))
  track_number_extrema_Lund <- numeric(ncol(Lund.methyl))
  for (i in 1L:ncol(Lund.methyl)) {
    
    are_na <- is.na(Lund.methyl[, i])
    rm_extremes <- !are_na & {Lund.methyl[, i] <= 0.98 & Lund.methyl[, i] >= 0.02}
    track_number_removed_Lund[i] <- sum(!rm_extremes)
    track_number_extrema_Lund[i] <- sum(!are_na & {
      Lund.methyl[, i] > 0.98 | Lund.methyl[, i] < 0.02})
    res <- tryCatch(suppressWarnings(BetaMixture(Lund.methyl[rm_extremes, i], K = 3, 
                                                 forever = 1500,
                                                 debug = FALSE,
                                                 epsilon = 1e-3, 
                                                 relative = 1e-3)),
                    error = function(e) {
                      message(i, " ", sum(is.na(Lund.methyl[, i])), " ", e$message)
                      stop("T")
                    })
  
    hist(res, main = paste("Tumor ", colnames(Lund.methyl)[i]))
    if (is.null(res)) next
    if (!res@converged) {
      message(i, " ", res@converged)
    }
    alpha[i, ] <- res@mle[1L, ]
    delta[i, ] <- res@mle[2L, ]
    z[[i]] <- res@Z
    phi[i, ] <- res@phi
  }
  names(z) <- colnames(Lund.methyl)
  rownames(phi) <- colnames(Lund.methyl)
  names(track_number_removed_Lund) <- colnames(Lund.methyl)
  
  ########################################################  
  ############# Saving results as R object ###############
  ########################################################
  
  Lund_output_filepath <- file.path(outdirpath, params$Lund_output_filename)
    
  
  # Collect old results
  if (params$report_updates) {
    ident.vec <- NULL
    if (file.exists(Lund_output_filepath)) {
      prev.results <- readRDS(Lund_output_filepath)
      
      # Compare all data objects
      identical.all <- all(c(
        identical(prev.results$alpha, alpha),
        identical(prev.results$delta, delta),
        identical(prev.results$Z, z),
        identical(prev.results$n_dropped_sites, track_number_removed_Lund),
        identical(prev.results$phi, phi)
      ))
      
      if (identical.all) {
        print('Results DO match previously saved results, resaving anyway')
      } else{
        print('Results do not match previously saved results')
        
        Lund_output_filepath <- paste(c(Lund_output_filepath, as.character(Sys.time())), collapse='.')
        sprintf('Will not override, saving results instead as %s', Lund_output_filepath)
      }
    }
  }
  
  # Save results in R data object just in case
  saveRDS(list(alpha = alpha, delta = delta, Z = z, 
               n_dropped_sites = track_number_removed_Lund,
               phi = phi), 
          file = file.path(outdirpath, params$Lund_output_filename))
  
  ########################################################
  ########## Output results as readable tables ###########
  ########################################################
  write.table(alpha, file.path(outdirpath, 'alpha.txt'), quote=F)
  write.table(delta, file.path(outdirpath, 'delta.txt'), quote=F)
  write.table(phi, file.path(outdirpath, 'phi.txt'), quote=F)
  
  modes <- (alpha - 1)/(alpha + delta - 2)
  write.table(modes, file.path(outdirpath, 'modes.txt'), quote = F)
  ########################################################
  ########################################################
  
  print(table(track_number_removed_Lund))
  print(table(track_number_extrema_Lund))
  sprintf('Total # extreme values removed: %d', sum(track_number_extrema_Lund))
}
```